---
title: "From a Taxonomy Table to an Interactive Sunburst Chart"
subtitle: "A step-by-step R tutorial (amplicon/metabarcoding data)"
author: "Giovanna Lanzavecchia"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Why this tutorial

Krona charts are a familiar sight in metagenomics: a multi-ring
interactive pie chart you can click into, level by level, to explore a
taxonomic community. They're generated by a standalone command-line tool
though, which means the data-wrangling step — going from "a feature
table and a taxonomy file" to "something a sunburst chart can actually
plot" — stays hidden.

This tutorial builds the same kind of chart **natively in R**, with
every step visible: parsing taxonomy strings, aggregating abundance up a
hierarchy, and rendering an interactive `plotly` sunburst you can export
as a single, self-contained HTML file — open it in any browser, no
server or R installation required on the viewer's end.

> **Demo data note:** the input files are simulated
> (`R/00_generate_demo_data.R`). Taxon names are real, standard fungal
> reference taxonomy; the read counts are synthetic. Swap in your own
> QIIME2/DADA2 output and everything below works unchanged — see the
> last section.

```{r libs}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(plotly)
library(htmlwidgets)

source("R/utils.R")
```

## 1. The data: a feature table + a taxonomy file

Amplicon pipelines (QIIME2, DADA2, mothur...) typically hand you two
separate files that share a `Feature_ID` (ASV/OTU) key:

-   a **feature table**: how many reads of each ASV were seen in each
    sample
-   a **taxonomy file**: what each ASV was classified as, usually as a
    single semicolon-delimited string
    (`d__Fungi;p__Ascomycota;c__...;o__...;f__...`)

```{r load-data}
feature_table <- read_csv("data/feature_table.csv", show_col_types = FALSE)
taxonomy_raw  <- read_csv("data/taxonomy.csv", show_col_types = FALSE)

feature_table
taxonomy_raw
```

## 2. Parsing the taxonomy string

That `Taxon` column is the first real obstacle: it's one string per ASV,
not the separate Kingdom/Phylum/Class/... columns we actually need. The
`d__` / `p__` / `c__` prefixes are the SILVA/QIIME2 rank convention.

```{r parse-taxonomy}
taxonomy <- parse_taxonomy(taxonomy_raw)
taxonomy
```

`parse_taxonomy()` (in `R/utils.R`) does two things: `tidyr::separate()`
on `;`, then strips the rank-prefix from each resulting column with a
small regex. Worth a look if you haven't split-and-cleaned a taxonomy
string before — it's a pattern you'll reuse constantly in metabarcoding
work.

## 3. From counts to a taxonomy-annotated long table

Now we reshape the feature table to long format and join in the
taxonomy, so every read count is tagged with its full lineage.

```{r long-format}
long_data <- feature_table %>%
  pivot_longer(-Feature_ID, names_to = "Sample", values_to = "Abundance") %>%
  left_join(taxonomy, by = "Feature_ID")

long_data
```

Multiple ASVs often share the same Family (or even Genus/Species) —
several distinct sequence variants can belong to the same named taxon.
For the sunburst we care about total abundance *per named taxon*, not
per ASV, so we aggregate:

```{r aggregate-family}
family_totals <- long_data %>%
  group_by(Phylum, Class, Order, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  arrange(desc(Abundance))

family_totals
```

## 4. The key idea behind a sunburst: values at every level

A sunburst plot needs a numeric value at **every node**, not just the
leaves — the Phylum ring's segment sizes are sums of their Class
children, which are sums of their Order children, and so on.
`build_sunburst_df()` does this generically for any number of rank
columns:

```{r show-helper}
print(build_sunburst_df)
```

```{r build-sunburst-df}
sun_df <- build_sunburst_df(
  family_totals,
  rank_cols = c("Phylum", "Class", "Order", "Family"),
  value_col = "Abundance"
)

sun_df <- color_by_top_level(sun_df)
sun_df
```

Each row is one ring segment: `id` uniquely identifies it (we build IDs
by joining rank names with `/`, so `"Ascomycota/Sordariomycetes"` is
unambiguous even if two phyla happened to share a class name), `parent`
points to its ring segment one level up, and `value` is its aggregated
abundance. This `id`/`label`/`parent`/`value` shape is exactly what
`plotly`'s sunburst trace expects.

## 5. Plotting it

```{r plot-sunburst}
fig <- plot_ly(
  sun_df,
  ids = ~id,
  labels = ~label,
  parents = ~parent,
  values = ~value,
  type = "sunburst",
  branchvalues = "total",
  marker = list(colors = ~color),
  hovertemplate = "<b>%{label}</b><br>Reads: %{value}<br>%{percentParent} of parent<extra></extra>"
)

fig
```

`branchvalues = "total"` tells plotly that a parent's value already *is*
the sum of its children (which is what we computed) rather than
something it should sum for us — get this backwards and every ring
balloons to the wrong size.

Click any ring segment to zoom in; click the centre to zoom back out.

## 6. Exporting a standalone, shareable HTML file

This is the part that replaces Krona's "just open the HTML" convenience:

```{r export-html}
dir.create("output", showWarnings = FALSE)
saveWidget(fig, "output/sunburst_demo.html", selfcontained = TRUE)
```

`output/sunburst_demo.html` is now a single file with the chart and all
its JavaScript baked in. Anyone can double-click it and get the
interactive chart in their browser — no R, no package installs, no
server. It's also what you'd point GitHub Pages at if you wanted a live
link instead of a downloadable file.

## 7. Going further: a Shiny version

A static sunburst is great for a report. If you want a
**sample-by-sample explorer**, wrap steps 3–5 in a Shiny app with a
dropdown — see [`app.R`](app.R) in this repo for a minimal version
(`shiny::runApp()` to launch it). The core plotting code doesn't change
at all; only a `selectInput()` for `Sample` and a `filter()` before the
aggregation step.

## Using this with your own data

1.  Export your feature table and taxonomy from QIIME2
    (`qiime tools export`) or your DADA2/phyloseq object as two CSVs
    shaped like `data/feature_table.csv` / `data/taxonomy.csv`.
2.  If your taxonomy strings use different rank prefixes or a different
    number of ranks, adjust the `into =` argument in `parse_taxonomy()`
    (`R/utils.R`).
3.  Everything from Section 3 onward works unchanged.
